Left-Linear Rewriting in Adhesive Categories

### 2024-07-10 concur/final.tex
minori
minima formattazione Def. 3.1.

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
introducing the standard notion of independence and 
a novel axiomatic notion of switchability, showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching rewriting systems,
a subclass of left-linear rewriting systems,
and we prove that the classical results for linear rewriting systems
are recovered for such a class.
Finally, in \S\ref{conclusions} we outline directions for future work.
%
%The appendices
% collect additional data,
Appendix~\ref{app:ade} recalls basic results on adhesive categories and
Appendix~\ref{app:fill}
presents \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds. Full proofs of our results are available in the extended version of the paper \cite{baldan2024full}.
\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$.
We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ if
$n<0$.

@@

%It can be seen that 
$\mathcal{M}$-adhesivity ensures the
uniqueness of the result of applying a rule to an object: two
direct derivations using the same match are abstraction equivalent
(see Proposition~C.1 of the extended version).

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  We say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  \emph{sequentially independent}
  if there is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes. In this case $(i_0,i_1)$ is called an  \emph{independence
    pair}.
  \[\xymatrix@R=16pt@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}
  \]
\end{definition}

>>>>

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
introducing the standard notion of independence and 
a novel axiomatic notion of switchability, showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching rewriting systems,
a subclass of left-linear rewriting systems,
and we prove that the classical results for linear rewriting systems
are recovered for such a class.
Finally, in \S\ref{conclusions} we outline directions for future work.
%
%The appendices
% collect additional data,
Appendix~\ref{app:ade} recalls basic results on adhesive categories and
Appendix~\ref{app:fill}
presents \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds. Appendix C of the
extended version provides the proofs
of our results.

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This  section recalls the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.
%
%Throughout the paper
Given a category $\X$ we will not distinguish
notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$.
We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of
all arrows, monos and regular monos of $\X$, respectively.  Given an
integer $n\in \mathbb{Z}$, $[0,n]$ denotes the set of natural numbers
less than or equal to $n$; in particular, $[0,n]=\emptyset$ if
$n<0$.

@@

%It can be seen that 
$\mathcal{M}$-adhesivity ensures the
uniqueness of the result of applying a rule: two
direct derivations using the same match are abstraction equivalent
(see Proposition~C.1 of the extended version).

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  We say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  \emph{sequentially independent}
  if there is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes
  \[\xymatrix@R=16pt@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}
  \]
 In this case $(i_0,i_1)$ is called an  \emph{independence pair}.
    \end{definition}
